I don't know, are you sure we want to do this?\\ Sea level adaptation decisions under uncertainty

### 2016-12-29 submission/paper.tex
updating paper

\begin{keypoints}
\item Decisions on adaptation measures need to take careful account of uncertanties
\item More careful models are needed for increased costs of the effects of sea level rise
\item Modelling local sea level rise is essential
\end{keypoints}

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, in the cost of no action. We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark. Different components of uncertainty are visualized. We show that failing to take uncertainty into account can increase the damage costs by an order of magnitude.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate. Especially as the potential impact of climate change on local sea level can yield numerous effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats (add REF). Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts. Severe inherent uncertainties persist along all parts of the processing chain from climate projections to adaptation assessments, and it is critical that decision-making appropriately accounts for this.

@@

We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark. The Norwegian city of Bergen is the capital of Hordaland County. The city center is located on Byfjorden, and is surrounded by mountains. It has the largest port in Nor way, both in terms of freight and passengers. The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}. Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become  a major problem in other parts of Bergen as well \citep{bergenreport}. Esbjerg, on the west coast of Jutland, has the second largest harbour in Denmark. It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.

Section \ref{sealevelproj} describes our approach to projecting sea level. In section \ref{decision_tools} we describe the decision framework focused on the timing of proactive adaptation decision. We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.

\section{Sea level projections}

\label{sealevelproj}

@@

We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as

\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t}\]
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as

\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]

where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.

@@

\section{Decision tools {\color{blue} (KdB, MD, TT)}}
\label{decision_tools}

\subsection{Timing of adaptation measures}

We consider adaptation decision making related to the timing of proactive adaptation measures. That is, the goal is to adapt to sea level rise before major damages occur. In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) \citep{Fankhauser&1999}.

We assume that damages related to sea level rise coincide with damages due to storm surges and that the accumulated annual damages are independent between years. More specifically, we model the distribution of annual damage, $F_{\textup{d}}$, by the three parameter Burr distribution \citep{Burr1942} with density
\begin{linenomath*}
  \begin{equation}\label{eq:Burr}
  f_{\textup{d}} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
  \end{equation}
\end{linenomath*}
for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter. The Burr distribution has a heavy upper tail and is commonly used to model damage loss, see e.g. \cite{Klugman&2012}.

Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d. values from the estimated damage distribution for $t_0 = 2015$, $\hat{F}_{\textup{d}, t_0}$.  By repeating this process $J$ times, we obtain an empirial damage distribution for each future year $t_i$ given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{d_{t_i}^{(j)} \leq x \}
  \]
\end{linenomath*}
for $i = 1, \ldots, 85$. Similarly, we obtain an empirial distribution of the total damage over the period $2016-2100$ by considering $d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} d_{t_i}^{(j)}$.

Now assume we are given a large sample of projections $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ for annual sea level rise anomalies compared to the 2015 value for the same time period. We model the relationship between the change in damage and change in sea level by a monotonic positive function $g(s| \boldsymbol \beta )$ with parameters $\boldsymbol \beta$, where $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$. We may further incorporate uncertainty in the shape of the function $g$ through the values of the parameters $\boldsymbol \beta$. An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and change in damage due to sea level rise is then given by
\begin{linenomath*}
  \begin{equation}\label{eq:FutureDamage}
    \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}.
  \end{equation}
\end{linenomath*}

The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures. In addition, we can incorporate an adaptation measure that protects against $K$ cm of increased sea level from year $t_k$ onwards. This results in a damage distribution given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i} (x) =
    \begin{cases}
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i < t_k \\
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i \geq t_k. \\
    \end{cases}
  \]
\end{linenomath*}

\begin{comment}
\cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$. If $r$ is the discount rate, the net present value damage, $D^0$, associated with this investment is
\begin{linenomath*}
\begin{equation}\label{eq:deterministic damage}
D^0 = C^0 + d_0^0 + \frac{d_1^0}{1+r} + \frac{d_2^0}{(1+r)^2} + \cdots  
\end{equation}
\end{linenomath*}
In comparison, postponing the adaptation investment to time period $n=1$ would lead to unmitigated damages in periods $0$ and $1$, and partially mitigated damages, $d_t^1$, thereafter. The delay would be preferable if
\begin{linenomath*}
\[
C^0 - \frac{C^1}{(1+r)} > (d_0^1 - d_0^0) + \frac{d_1^1 - d_1^0}{1+r} + \frac{d_2^1 - d_2^0}{(1+r)^2} + \cdots
\]
\end{linenomath*}
Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay. In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$). In this case, the comparison is between the expected return $r$ earned on the capital while implementation is delayed and one additional time period of unmitigated damage,
\begin{linenomath*}
  \[
  r C > d_1^1 - d_1^0.
  \]
  \end{linenomath*}
\end{comment}

\subsection{Limitations of the decision framework}

@@

\subsection{Data {\color{blue} (PG)}}
The historical global mean temperature series is obtained from \citet{giss}. Climate projections of global mean temperature are from the fifth climate model intercomparison project, CMIP5 \citep{cmip5}. The global mean sea level series is obtained from \citet{csiro}. We use local tide gauge data from the Permanent Service for Mean Sea Level, UK, which is the worldwide repository for national sea level data. Glacial isostatic adjustment for Bergen is obtained from \citet{Simpson2014}, and for Esbjerg in personal communication from Peter Thejll at the Danish Meteorological Institute.

The Bergen monthly series is missing data for 62 months, including all of the years 1942--43. To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages. For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other Norwegian stations corrected by the average difference.

@@

We estimate a damage distribution for Bergen in 2015 using reported annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}). The NPP data are aggregated to a county level, and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion. Also, some of this discussion should probably be moved to the section on data.}. In order to improve the parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland. Rogaland has a similar geography to Hordaland and a slightly smaller population with its main city, Stavanger, also located on the coast. The damage data prior to 2015 is adjusted to the 2015 level using the consumer price index for Norway. After the adjustment, we assume stationarity over the period and fit a single Burr distribution to the data, cf. \eqref{eq:Burr}.  
\subsection{Sea level rise in Bergen and Esbjerg}

Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data. The glacial isostatic adjustment is 0.26 (0.07) cm/yr. The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17). The regression slope is 1.30 (0.12).
\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=\linewidth]{bergenfit.png}
\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}. The straight line is the time series regression line.}
\label{fig:bergenobs}
\end{center}
\end{figure}

@@

For modelling the relationship between the change in damage and change in sea level we use the results of \cite{Hallegatte&2013} for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseille, Naples, Porto, Rotterdam and Stockholm. \cite{Hallegatte&2013} estimate the mean annual flood damage in these cities in 2050 under no sea level change, 20 cm increase and 40 cm increase in sea level. We standardize their results for each city and consider the relative change for 20 and 40 cm increase. We then perform a linear extrapolation to obtain estimated relative changes in damage for a large interval of changes in sea level rise, see Figure~\ref{fig:ChangeInDamage}. Finally, the function $g$ is obtained by sampling from this ensemble of functions with all 15 ensemble members considered equally probable.

@@

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.5\linewidth]{unc.pdf}
\caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5. The black vertical line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen. } 
\label{fig:unc}
\end{center}
\end{figure}

>>>>

\begin{keypoints}
\item Decisions on adaptation measures need to take careful account of uncertainties
\item More careful models are needed for increased costs of the effects of sea level rise
\item Modeling local sea level rise is essential
\end{keypoints}

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action. We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark. Different components of uncertainty are visualized. We show that failing to take uncertainty into account can increase the projected damage costs by an order of magnitude.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate. In particular, potential changes in local sea level can yield numerous effects such as frequent flooding, inundation and back flow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats. Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts. Severe inherent uncertainties persist along all parts of the processing chain from climate projections to adaptation assessments, and it is critical that decision-making appropriately accounts for this.

@@

We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark. The Norwegian city of Bergen is the capital of Hordaland County. The city center is located on Byfjorden, and is surrounded by mountains. It has the largest port in Norway, both in terms of freight and passengers. The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}. Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.

The municipality of Bergen has, in cooperation with private actors, analyzed several possible adaptation measures against sea level rise. The measures range from an outer barrier that would protect the entire metropolitan area to various protection measures of limited areas in the inner harbor \citep{bergenreport}. While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated. We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs and the effects of sea level rise on changes in damage costs.

Esbjerg, on the west coast of Jutland, has the second largest harbor in Denmark. It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.

The remainder of the paper is organized as follows. Section \ref{sealevelproj} describes our approach to projecting sea level. In section \ref{decision_tools_PartI} we describe the decision framework focused on the timing of proactive adaptation decision. We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.

\section{Sea level projections}
\label{sealevelproj}

We project local sea level changes by Modeling two processes:
\begin{enumerate}
\item The relationship between global temperature and global sea level;
\item The relationship between global sea level and local sea level.
\end{enumerate}

@@

We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}). 
For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
\begin{linenomath*}
\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
\end{linenomath*}
where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
\begin{linenomath*}
\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
\end{linenomath*}
where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.

@@

\section{Timing of adaptation measures}
\label{decision_tools_PartI}

To assess the optimal timing of an adaptation measure, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options. The probabilistic distribution is constructed by considering uncertainty
\begin{itemize}
\item in the local sea level projections;
\item in the annual damage costs;
\item in the effect of changes in sea level on the annual damage costs. 
\end{itemize}

\subsection{Annual damage costs}

We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
\begin{linenomath*}
  \begin{equation}\label{eq:Burr}
  f_{\textup{d}, t_0} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
  \end{equation}
\end{linenomath*}
for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter. The Burr distribution has a heavy upper tail and is commonly used to model damage loss, see e.g. \cite{Klugman&2012}. The parameters of the distribution are estimated using historical data for annual storm surge damage. Data prior to 2015 are adjusted to the 2015 level using the consumer price index. After adjustment, we assume stationarity over the period and independence between years.

Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d. values from the estimated distribution $\hat{F}_{\textup{d}, t_0}$.  By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l}) } \leq x \Bigg\}
  \]
\end{linenomath*}
for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
\begin{linenomath*}
  \[
  d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})}
  \]
\end{linenomath*}
and similarly for the cumulative damage.

\subsection{Effect of changes in sea level}

We assume that changes in sea level have a multiplicative effect on the annual damage cost. That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
\begin{linenomath*}
  \[
  g(s_{t_i} | \boldsymbol \beta ) d_{t_i},
  \]
\end{linenomath*}
where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$. \cite{Hallegatte&2013} estimate a similar effect function valid in 2050 for $s \in \{0,20,40\}$ for 136 coastal cities. Here, we use their results for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseilles, Naples, Porto, Rotterdam and Stockholm.  To obtain a city-specific effect function for a large range of sea level anomalies we employ a linear extrapolation as shown in Figure~\ref{fig:EffectFct}. We then obtain a sample of effect functions $\{g(\cdot|\boldsymbol \beta^{(j)})\}_{j=1}^J$ by sampling with replacement from this ensemble of trajectories with all 15 ensemble members considered equally probable.

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.5\linewidth]{DecisionAnalysisBergenPartII.pdf}
\caption{Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines. The median change and the corresponding extrapolation are indicated in red.}
\label{fig:EffectFct}
\end{center}
\end{figure}

Let further $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ denote a sample of projections for annual sea level anomalies compared to the 2015 value. An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
\begin{linenomath*}
  \begin{equation}\label{eq:FutureDamage}
    \hat{F}_{\textup{d}, t_i}^\textup{s} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}.
  \end{equation}
\end{linenomath*}

The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures. In addition, we can incorporate an adaptation measure of cost $C$ that protects against $K$ cm of increased sea level from year $t_k$ onward. This results in a damage distribution given by
\begin{linenomath*}
  \[
  \hat{F}_{\textup{d}, t_i}^{\textup{s}, \textup{a}_k} (x) =
    \begin{cases}
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i < t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} + C}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i = t_k \\[2ex]
      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i > t_k. \\
    \end{cases}
  \]
\end{linenomath*}

\subsection{Limitations of the decision framework}
The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs. In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise. However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible. Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level. Additionally, we have not specifically accounted for potential changes in storm surge patterns.

\section{Danish decision framework}

@@

\subsection{Data}
The historical global mean temperature series is obtained from \citet{giss}. Climate projections of global mean temperature are from the fifth climate model intercomparison project, CMIP5 \citep{cmip5}. The global mean sea level series is obtained from \citet{csiro}. We use local tide gauge data from the Permanent Service for Mean Sea Level, UK, which is the worldwide repository for national sea level data. Glacial isostatic adjustment for Bergen is obtained from \citet{Simpson2014}, and for Esbjerg in personal communication from Peter Thejll at the Danish Meteorological Institute.

The Bergen monthly series is missing data for 62 months, including all of the years 1942--43. To deal with occasional short stretches of missing data (at most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages. For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other Norwegian stations corrected by the average difference.

@@

Annual damage costs for the Bergen case study are obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}). The NPP data are available for the period 1980-2015 and are aggregated to a county level. For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics. We use a dicsount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf. Section 5.8 of \citet{DiscountRate}).

\subsection{Sea level rise in Bergen and Esbjerg}

Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data. The glacial isostatic adjustment is 0.26 (0.07) cm/yr. The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17). The regression slope is 1.30 (0.12).

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.75\linewidth]{bergenfit_edit.png}
\caption{ The left figure shows raw (black) and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}. The straight line is the time series regression line.}
\label{fig:bergenobs}
\end{center}
\end{figure}

@@

@@

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.5\linewidth]{unc.pdf}
\caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5. The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen. } 
\label{fig:unc}
\end{center}
\end{figure}
